Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-09-06 Manuscript/00-main.Rmd
Manuscript revision

---
title: Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: Byron C. Jaeger
  num: b 
- name: AKM Fazlur Rahman
  num: a   
- name: D. Leann Long
  num: a  
- name: Nengjun Yi
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
- num: b
  org: Department of Biostatistics and Data Science, Wake Forest University, Winston-Salem, USA
corres: "Nengjun Yi, epartment of Biostatistics, University of Alabama at Birmingham, Birmingham, USA. \\email{nyi@uab.edu}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo et al.
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "Spike-and-slab priors have been extensively studied for variable selection, and recently been extended to function selection in generalized additive models (GAMs). Most of the previous works on fitting spike-and-slab GAM relies on Markov chain Monte Carlo (MCMC) algorithms to identify the promising subsets of functions. The computational burden of MCMC makes these models less feasible, especially in high-dimensional setting. Besides, most of the current spike-and-slab GAM consider an “all-in-all-out” approach for function selection, rendering less flexibility to answer if non-linear effects are necessary in the model. We propose Bayesian hierarchical generalized additive models for function selection and prediction. The proposed models employ spike-and-slab priors, mixture normal and mixture double exponential, for smooth and sparse solution. Two EM based algorithms, EM-IWLS and EM-Coordinate Descent, are developed for accelerated computation. Simulation studies and real data analysis demonstrated improved performance against the state-of-art models, mgcv and COSSO. The software implementation of the proposed methods is freely available via the R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

@@

library(knitr)
library(targets)
library(here)
library(xtable)

@@

tar_load(ful_dat)
tar_load(tbl_one)
tar_load(bamlasso_tune)
tar_load(bai_cv_res)
```

```{r child=here(intro_path)}
```

```{r child=here(method_path)}
```

```{r child=here(sim_path)}
```

```{r child=here(real_data_path)}
```

```{r child=here(disc_path)}
```

```{r plot-ref, fig.cap = "Fancy Caption\\label{fig:plot}", eval = F}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
```

<!-- ...and can be referenced (Figure \ref{fig:plot}) by including the `\\label{}` tag in the `fig.cap` attribute of the R chunk: `fig.cap = "Fancy Caption\\label{fig:plot}"`. It is a quirky hack at the moment, see [here](https://github.com/yihui/knitr/issues/323). -->

<!-- Analogously, use Rmarkdown to produce tables as usual: -->

```{r, results = "asis", eval = F}
if (!require("xtable")) install.packages("xtable")
xt <- xtable(head(cars), caption = "A table", label = "tab:table")
print(xt, comment = FALSE)
```

>>>>

---
title: Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data
author:
- name: Boyi Guo*
  num: a
- name: Byron C. Jaeger
  num: b 
- name: AKM Fazlur Rahman
  num: a   
- name: D. Leann Long
  num: a  
- name: Nengjun Yi
  num: a
address:
- num: a
  org: Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA
- num: b
  org: Department of Biostatistics and Data Science, Wake Forest University, Winston-Salem, USA
corres: "Nengjun Yi, Department of Biostatistics, University of Alabama at Birmingham, Birmingham, USA. \\email{nyi@uab.edu}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Guo et al.
articletype: Research article
received: 2021-01-01
revised: 2021-02-01
accepted: 2021-03-01
abstract: "There are many proposals that extend the classical generalized additive models (GAMs) to accommodate the high-dimensional (p>>n) problem using group penalties and priors. These models rarely smooth the nonlinear effect properly because of the sparse regularization. It leads to excess shrinkage on coefficients and inaccurate prediction. Besides, most of these GAMs consider an “all-in-all-out” approach for function selection, rendering them difficult to answer if nonlinear effects are necessary. While some Bayesian models can address the shortcomings, using Markov chain Monte Carlo (MCMC) algorithms for model fitting creates a new challenge, scalability. Hence, we propose Bayesian hierarchical generalized additive models as a solution. For smooth and sparse solutions, the proposed models employ bi-part spike-and-slab spline priors, including mixture double exponential and mixture normal priors. Two optimization-based algorithms, EM-Coordinate Descent and EM-Iterative Weighted Least Squares, are developed for expedited computation. Simulation studies and metabolomics data analyses demonstrate improved performance against state-of-art models, mgcv, COSSO and sparse Bayesian GAM. The software implementation of the proposed models is freely available via an R package BHAM."
keywords: Spike-and-Slab Priors; High-Dimensional Data; Generalized Additive Models; EM-IWLS; EM-Coordinate Decent
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    keep_tex: TRUE
header-includes: 
  - \usepackage{tikz}
  - \usepackage{pgfplots}
---

@@

library(knitr)
library(targets)
# library(here)
library(xtable)

@@

tar_load(WLM_train_dat)
# tar_load(tbl_one)
tar_load(ECB_bamlasso_cv)
tar_load(ECB_SBGAM_cv)
tar_load(WLM_bamlasso_cv)
tar_load(WLM_SBGAM_cv)
```

```{r child="01-Intro.Rmd"}
```

```{r child="02-Method.Rmd"}
```

```{r child="03-Simulation.Rmd"}
```

```{r child="04-Real_Data.Rmd"}
```

```{r child="05-Discussion.Rmd"}
```

### 2021-09-06 Manuscript/01-Intro.Rmd
Manuscript revision

Many modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see Mallick and Yi [@Mallick2013] for review. These models are built upon a restrictive and unrealistic assumption, linearity. In classical statistical modelling, many modeling strategies and models are proposed to relax the linearity assumption with various degrees of complexity. For example, variable categorization is a simple and common practice in epidemiology, but suffers from power and interpretation issues. More complex models to address nonlinear effects include random forest and other so-called "black box" models [@Breiman2001]. These models are not particularly attractive in biomedical studies, as they lack the transparency to understand the etiology of diseases. In addition, how to generalized these models to the high-dimensional setting remains unclear.

@@

Semi-parametric regression models are an appropriate alternative to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be expressed as the summation of smooth functions, $B_j, j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
$$
where $g^{-1}$ is the inverse function of a monotonic link function. The smooth functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

There exists some proposals to generalized the classical GAM to accommodate high-dimensional applications. The regularized models that are primarily designed for linear models are used to fit GAMs by accounting for the group structure introduced when expanding smoothing functions.
<!-- These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients:  -->
Ravikumar et al. [@Ravikumar2009] extended the grouped lasso [@Yuan2006] to additive models; Huang et al. [@Huang2010] further developed adaptive grouped lasso for additive models; Wang et al. [@Wang2007] and Xue [@Xue2009] respectively applied grouped SCAD penalty [@Fan2001] to the additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed, where the application on AMs are by-products. Bai et al. [@Bai2020] was the first to apply group spike-and-slab lasso prior to Gaussian additive models(AMs) using a fast optimization algorithm, and further generalized the framework to GAMs[@Bai2021]. As these methods focus on addressing the sparsity, it is very likely to overly penalize the basis function coefficients and produce inaccurate predictions and curve interpolation, particularly when complex signals are assumed and large number of knots are used. [@Scheipl2013] In addition, the proposed methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for bi-level selection. Scheipl et al. [@Scheipl2012] proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma prior, that address the previous challenges. But the model fitting relies on computational intensive MCMC algorithmic and creates scalability concern.
 <!-- specifically for semi-parametric regression using MCMC. The normal-mixture-of-inverse gamma prior is mixture of t-distribution with an additional coefficients mixing vector, where the vector follows a mixture of two Gaussian distribution. -->
It would be of special interest to develop a fast, flexible and accurate generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we incorporate smoothness penalties in the model via re-parameterization of the smoothing function to avoid overly shrinking basis function coefficients. Smoothness penalties are commonly implemented in GAMs through smoothing regression splines. They are quadratic norms of the coefficients and allow locally adaptive penalties on each smoothing function. The smoothness penalty conditioning on a smoothness penalty parameter $\lambda_j$ is a function of the integration of the second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known smoothness penalty matrix, and $\bs \beta_j$ is the basis function coefficients. Smoothness penalties are also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothness penalty allows the separation of the linear space of smooth functions, also know as the null space, from the nonlinear space. We then impose a new bi-part spike-and-slab spline prior on the smooth functions for bi-level selection such that the linear and nonlinear spaces of smooth functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. In addition, two fast computing optimization-based algorithms are developed for high and ultrahigh-dimensional[@Fan2009] settings respectively.

@@

The rest of the paper are arranged as the followings: in Chapter 2 we demonstrate the Bayesian hierarchical generalized additive model, where we provide the setup of the proposed model and describe the two fast-fitting algorithms. In Chapter 3, we compare the proposed via monte carlo simulation. .Application to real-world datasets demonstrates the usefulness of the proposed methodologies in Chpater 4. Conclusion and discussions are made in Chapter 5. ^[TODO: rewrite this sentence.]

@@

<!-- To the best of our knowledge, there are few GAM models proposed for high dimensional application. In common practice, model matrix of spline terms are generated using base functions of preference, and then feed to a penalized model. Without proper handle of the grouping nature of bases functions, the estimation of the smooth function is less accurate. @Bai2020Spline shows inferior performance of such practice compared to using an ad hoc high-dimensional additive model. Among the specialized high-dimensional spline methods, @Ravikumar2009 extended the original backfitting algorithm [@Hastie1987] to accommodate the high-dimensional setting. The sparsity and smoothness was handled separately with a soft threshold. The authors also demonstrated its similarity to Group Lasso. @Meier2009 proposed an a sparsity-smoothness penalty, -->

@@

<!-- Among all the choices of smoothing function, one particular family stood out, splines. Splines are piece-wise polynomial functions to interpolate nonlinear signals. Most commonly used spline is cubic spline, thanks to its smoothness property. [@Wood2017, PP.197] Early spline research focused on regression splines, where number and location of knots are pre-determined and ordinary least squares is used to estimate the smooth function. However, the specification of knots is not a trivial task, as the the positioning of knots are influential on the performance of the model. As an improvement, researchers proposed smoothing splines where as many knots are used and later shrunk using penalized algorithms. The smoothing spline concept was first introduced [@OSullivan1986; @OSullivan1988], and later generalized and made famous by @Eilers1996, called P-spline. Independently, @Ruppert1999 proposed similar idea. Later, Bayesian GAMs were introduced as an alternative to address measurement error in covariates [@Berry2002] or local adaptive smoothing [@Lang2004]. -->

>>>>

Many modern biomedical research, e.g. sequencing data analysis, electric health record data analysis, require special treatment of high-dimensionality, commonly known as $p >> n$ problem. There is extensive literature on high-dimensional linear models via penalized models or Bayesian hierarchical models, see Mallick and Yi [@Mallick2013] for review. These models are built upon a restrictive and unrealistic assumption, linearity. In classical statistical modelling, many strategies and models are proposed to relax the linearity assumption with various degrees of complexity. For example, variable categorization is a simple and common practice in epidemiology, but suffers from power and interpretation issues. More complex models to address nonlinear effects include random forest and other so-called "black box" models [@Breiman2001]. These models are not particularly attractive in biomedical studies, as they lack the transparency to understand the etiology of diseases. In addition, how to generalized these models to the high-dimensional setting remains unclear.

@@

Nonparametric regression models are appropriate alternatives to the "black-box" models thanks to their balance between model flexibility and interpretability. Among those, generalized additive models (GAMs), proposed in the seminal work of Hastie and Tibshirani [@Hastie1987], grew to be one of the most popular modeling tools. In a GAM, the response variable, $Y$, which is assumed to follow some exponential family distribution $EF(\mu, \phi)$ with mean $\mu$ and dispersion $\phi$, can be expressed as the summation of smoothing functions, $B_j, j = 1, \dots, p$, of a given $p$-dimensional vector of covariates $\bs x$, written as 
$$
 E(Y|\bs x) = g^{-1}(a + \sum\limits^p_{j=1}B_j(x_j)),
$$
where $g^{-1}$ is the inverse of a monotonic link function. The smoothing functions can take many forms and are estimated using a pseudo-weighted version of the backfitting algorithm [@Breiman1985]. Nevertheless, the classical GAMs cannot fulfill the increasing analytic demands for high-dimensional data analysis in biomedical studies.

There exists some proposals to generalize the classical GAM to accommodate high-dimensional applications. The regularized models, branching out from group regularized linear models, are used to fit GAMs by accounting for the structure introduced when expanding smoothing functions.
<!-- These models assume sparsity among the predictors, i.e. most predictors have small or no effects on the outcome. Further development focuses on addressing the grouping structure among the predictors, where AMs and GAMs can be considered special cases. Various penalties are imposed on the basis function coefficients:  -->
Ravikumar et al. [@Ravikumar2009] extended the grouped lasso [@Yuan2006] to additive models (AMs); Huang et al. [@Huang2010] further developed adaptive grouped lasso for additive models; Wang et al. [@Wang2007] and Xue [@Xue2009] respectively applied grouped SCAD penalty [@Fan2001] to additive models. Recent Bayesian hierarchical models are also used in the context of high-dimensional additive models. Various group spike-and-slab priors combining with computationally intensive Markov chain Monte Carlo (MCMC) algorithms [@Xu2015; @Yang2020] are proposed, where the application on AMs are by-products. Bai et al. [@Bai2020] was the first to apply group spike-and-slab lasso prior to Gaussian AMs using a fast optimization algorithm, and further generalized the framework to GAMs[@Bai2021]. As these methods focus on addressing the sparsity, it is very likely to overly penalize the basis function coefficients and produce inaccurate predictions and curve interpolation, particularly when complex signals are assumed and large number of knots are used. [@Scheipl2013] In addition, these methods adapt an 'all-in-all-out' strategy, i.e. either including or excluding the variable completely, rendering no space for bi-level selection. Scheipl et al. [@Scheipl2012] proposed an ad hoc spike-and-slab structure prior, normal-mixture-of-inverse gamma prior, that address the previous challenges. But the model fitting relies on computational intensive MCMC algorithmic and creates scalability concern. It would be of special interest to develop a fast, flexible and accurate generalized additive model framework.

To address these challenges, we propose a novel Bayesian hierarchical generalized additive model (BHAM) for high dimensional data analysis. Specifically, we incorporate smoothing penalties in the model via re-parameterization of the smoothing function to avoid overly shrinking basis function coefficients. Smoothing penalties are commonly implemented in the classical GAMs through smoothing regression splines. They are quadratic norms of the coefficients and allow locally adaptive penalties on each smoothing function. A smoothing penalty conditioning on a smoothing parameter $\lambda_j$ is a function of the integration of the second derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{\prime\prime}_j(x)dx = \lambda_j \bs \beta_j^T \bs S_j \bs \beta_j ,
\end{equation}
where $\bs S_j$ is a known smoothing penalty matrix, and $\bs \beta_j$ are the basis function coefficients. Smoothing penalties re also previously used in the spike-and-slab GAM [@Scheipl2012] and the sparsity-smoothness penalty [@Meier2009]. Moreover, incorporating the smoothing penalty allows the separation of the linear space of smoothing functions, also know as the null space, from the nonlinear space. We then impose a new bi-part spike-and-slab spline prior on the smoothing functions for bi-level selection such that the linear and nonlinear spaces of smoothing functions can be selected separately. The prior setup encourages a flexible solution, rending one of three possibilities: no effect, only linear effect, or nonlinear effect of a predictor. In addition, two fast computing optimization-based algorithms are developed for high and ultrahigh-dimensional[@Fan2009] settings respectively.

@@

In Section \ref{sec:BHAM}, we establish the Bayesian hierarchical generalized additive model, introduce the proposed spike-and-slab spline priors, and describe the two fast-fitting algorithms. In Section \ref{sec:sim}, we compare the proposed framework to state-of-art models, mgcv, COSSO and sparse Bayesian GAM via monte carlo simulation studies. Analyses of two metabolomics datasets using the proposed model are presented in Section \ref{sec:real_data}. Conclusion and discussions are given in Section \ref{sec:concl}.

@@

<!-- To the best of our knowledge, there are few GAM models proposed for high dimensional application. In common practice, model matrix of spline terms are generated using base functions of preference, and then feed to a penalized model. Without proper handle of the grouping nature of bases functions, the estimation of the smoothing function is less accurate. @Bai2020Spline shows inferior performance of such practice compared to using an ad hoc high-dimensional additive model. Among the specialized high-dimensional spline methods, @Ravikumar2009 extended the original backfitting algorithm [@Hastie1987] to accommodate the high-dimensional setting. The sparsity and smoothness was handled separately with a soft threshold. The authors also demonstrated its similarity to Group Lasso. @Meier2009 proposed an a sparsity-smoothness penalty, -->

@@

<!-- Among all the choices of smoothing function, one particular family stood out, splines. Splines are piece-wise polynomial functions to interpolate nonlinear signals. Most commonly used spline is cubic spline, thanks to its smoothness property. [@Wood2017, PP.197] Early spline research focused on regression splines, where number and location of knots are pre-determined and ordinary least squares is used to estimate the smoothing function. However, the specification of knots is not a trivial task, as the the positioning of knots are influential on the performance of the model. As an improvement, researchers proposed smoothing splines where as many knots are used and later shrunk using penalized algorithms. The smoothing spline concept was first introduced [@OSullivan1986; @OSullivan1988], and later generalized and made famous by @Eilers1996, called P-spline. Independently, @Ruppert1999 proposed similar idea. Later, Bayesian GAMs were introduced as an alternative to address measurement error in covariates [@Berry2002] or local adaptive smoothing [@Lang2004]. -->

### 2021-09-06 Manuscript/02-Method.Rmd
Manuscript revision

The re-parameterization sets up the foundation of the proposed model, which has three-fold benefits. First of all, as previously mentioned, the re-parameterization integrated the smoothness penalty into the design matrix, and hence the penalty can be expressed in the posterior density function implicitly in addition to the sparse penalty for variable and function selection. Secondly, the eigendecomposition of the smooth penalty isolates the linear space of a smooth function from the non-linear space, improving the feasibility of bi-level variable/function selection and estimation inference. The eigendecomposition also introduced the linearly independent column spaces of the smooth function matrix, which allows imposing independent priors on the coefficients. This reduces the computational complexity that multivariate priors induces, and greatly broadens the choices of priors and further model choices. Last but not least, the scaling step of $\bs X^\tp$ further simplifies the choice of penalty priors, from column-specific penalties to a unified penalty. This step is also necessary for the proposed spike-and-slab spline prior. Conventionally, the spike-and-slab priors require scaling of predictors ^[TODO: insert a citation.]

@@

where $\gamma^0_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}^{K_j}$ are latent indicator variables, indicating if the model includes the linear effect and the non-linear effect of the $j$th variable respectively. ; $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}. To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], where the $\bs \beta^0_j$ and $\bs \beta^\tp_j$ shares the same indicator variables $\gamma_j^0$, $\gamma_j^\tp$ respectively. It is possible to relax the assumption such that each coefficient $\beta$ to have its own latent indicator $\gamma$, but at the cost ofcomplicating the bi-level function selection. Conversely, it is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and non-linear effect, i.e. $\gamma_j = \gamma^0_j = \gamma^\tp_j$. This converges^[Is "becomes" better?] to the SB-GAM [@Bai2021].

@@

<!-- The main difference between Bai's work and the proposed models are two-folded: 1) the proposed models are feasible to answer the question if the non-linear effects are necessary for the predictors, while Bai's work can't due to the "all-in-all-out" approach; 2) the proposed models can provide uncertainly measures in addition to the point estimates of the coefficients, while Bai's work can't not due to the model fitting algorithm. One of the exception is Bai (in progress), who a grouped spike-and-slab LASSO prior [@Rockova2018] to address the grouping structure of spline bases.  -->

>>>>

The re-parameterization sets up the foundation of the proposed model, which has three-fold benefits. First of all, as previously mentioned, the re-parameterization integrated the smoothness penalty into the design matrix, and hence the penalty can be expressed in the posterior density function implicitly in addition to the sparse penalty for variable and function selection. Secondly, the eigendecomposition of the smooth penalty isolates the linear space of a smooth function from the nonlinear space, improving the feasibility of bi-level variable/function selection and estimation inference. The eigendecomposition also introduced the linearly independent column spaces of the smooth function matrix, which allows imposing independent priors on the coefficients. This reduces the computational complexity that multivariate priors induces, and greatly broadens the choices of priors and further model choices. Last but not least, the scaling step of $\bs X^\tp$ further simplifies the choice of penalty priors, from column-specific penalties to a unified penalty. This step is also necessary for the proposed spike-and-slab spline prior. Conventionally, the spike-and-slab priors require scaling of predictors ^[TODO: insert a citation.]

@@

where $\gamma^0_{j}\in\{0,1\}$ and $\gamma^\tp_{j}\in \{0,1\}^{K_j}$ are latent indicator variables, indicating if the model includes the linear effect and the nonlinear effect of the $j$th variable respectively. ; $s_0$ and $s_1$ are scale parameters, assuming $0 < s_0 < s_1$ and given. These scale parameters $s_0$ and $s_1$ can be treated as tuning parameters and optimized via cross-validation. A discussion of how to choose the scale parameters comes in Section \ref{sec:tune}. To note, this prior differs from previous group spike-and-slab lasso priors [@Tang2018; @Tang2019], where the $\bs \beta^0_j$ and $\bs \beta^\tp_j$ shares the same indicator variables $\gamma_j^0$, $\gamma_j^\tp$ respectively. It is possible to relax the assumption such that each coefficient $\beta$ to have its own latent indicator $\gamma$, but at the cost ofcomplicating the bi-level function selection. Conversely, it is possible to add a more restrictive assumption on the priors, assuming that one indicator variable decides the inclusion of both the linear effect and nonlinear effect, i.e. $\gamma_j = \gamma^0_j = \gamma^\tp_j$. This converges^[Is "becomes" better?] to the SB-GAM [@Bai2021].

@@

<!-- The main difference between Bai's work and the proposed models are two-folded: 1) the proposed models are feasible to answer the question if the nonlinear effects are necessary for the predictors, while Bai's work can't due to the "all-in-all-out" approach; 2) the proposed models can provide uncertainly measures in addition to the point estimates of the coefficients, while Bai's work can't not due to the model fitting algorithm. One of the exception is Bai (in progress), who a grouped spike-and-slab LASSO prior [@Rockova2018] to address the grouping structure of spline bases.  -->

### 2021-09-06 Manuscript/03-Simulation.Rmd
Manuscript revision

# Simulation Study

In this section, we compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for non-linear effect interpolation and prediction. SB-GAM is the first spike-and-slab lasso GAM. We implemented COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \text{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence used the default choices. Both mgcv and proposed models allow customized smoothing functions and we chose the cubic regression spline. We controlled the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. 5-fold cross-validation were used for COSSO models, SB-GAM and the proposed models for tuning parameter selection based on the default selection criteria. 20 default candidates were examined for SB-GAM and the proposed models which allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].

>>>>

# Simulation Study
\label{sec:sim}

In this section, we compare the performance of the proposed models to four alternative models: component selection and smoothing operator (COSSO) [@Zhang2006GAM], adaptive COSSO [@Storlie2011], generalized additive models with automatic smoothing [@Wood2011], SB-GAM [@Bai2021]. COSSO is one of the earliest smoothing spline models that consider sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using adaptive weight for penalties such that the penalty of each functional component are different for extra flexibility. Generalized additive models with automatic smoothing, hereafter \textit{mgcv}, is one of the most popular models for nonlinear effect interpolation and prediction. SB-GAM is the first spike-and-slab lasso GAM. We implemented COSSO and adaptive COSSO with R package \texttt{cosso 2.1-1}, generalized additive models with automatic smoothing with R package \texttt{mgcv 1.8-31}, SB-GAM with R package \text{sparseGAM 1.0}. COSSO models and SB-GAM do not provide flexibility to define smoothing functions, and hence used the default choices. Both mgcv and proposed models allow customized smoothing functions and we chose the cubic regression spline. We controlled the dimensionality of each smoothing function, 10 bases, for all different choices of smoothing functions. 5-fold cross-validation were used for COSSO models, SB-GAM and the proposed models for tuning parameter selection based on the default selection criteria. 20 default candidates were examined for SB-GAM and the proposed models which allow user-specification of tuning candidates. All computation was conducted on a high-performance 64-bit Linux platform with 48 cores of 2.70GHz eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R 3.6.2 [@R].

### 2021-09-06 Manuscript/04-Real_Data.Rmd
Manuscript revision

# Metabolomics Data Analysis

@@

```{r}

bam_mdl_res <- bamlasso_tune %>% 
  arrange(desc(auc)) %>% 
  slice(1) %>% 
  select(-s0)

rbind(
  'BHAM-SSL' = bam_mdl_res,
  'SB-GAM' = bai_cv_res %>% t %>%  data.frame()
) %>% 
  flextable() %>% 
  flextable::autofit() %>% 
  flextable::set_caption(caption = "Measures of optimal BHAM-SSL and SB-GAM models for Emory Cardiovascular Biobank by 10-fold cross-validation\\label{tab:ECB_res}")
```

## Weight Loss Maintenance Cohort 
We use the proposed models BHAM-SSL to analyze metabolomics data from a recently published study ^[TODO: add citation:https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0240764] on the association between metabolic biomarkers and weight loss. The data contained 765 metaboliltes in baseline plasma collected from three separate studies, profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation were performed during metabolites data preparation. The outcome of interest was percent change in insulin resistance, and hence modeled using a Gaussian model adjusting for covariates including age, sex, race, baseline clinical triglycerides, and percent change in weight over the follow-up time period. 731 individuals were included in the initial studies, and reduced to `r nrow(ful_dat)` after removing missing values. The descriptive statistics were tabulated in Table \ref{tab:tbl_one}.

```{r}
tbl_one %>% 
  as_flex_table() %>% 
  flextable::autofit() %>% 
  flextable::set_caption(caption = "The descriptive statistics of the sample in the metabolomics data analysis stratified by studies\\label{tab:tbl_one}")
```

>>>>

# Metabolomics Data Analysis
\label{sec:real_data}

@@

```{r}
rbind(
  'BHAM-SSL' = ECB_bamlasso_cv %>% slice_max(auc) %>% select(-s0),
  'SB-GAM' = ECB_SBGAM_cv %>%  slice_max(auc) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>% 
  flextable() %>% 
  flextable::autofit() %>% 
  flextable::set_caption(caption = "Measures of optimal BHAM-SSL and SB-GAM models for Emory Cardiovascular Biobank by 10-fold cross-validation\\label{tab:ECB_res}")
```

## Weight Loss Maintenance Cohort 
We use the proposed models BHAM-SSL to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publically available [@Bihlmeyer2021_data]. In this analysis, we primarily focused on the analysis of one of the three studies included, weight loss maintenance (WLM) cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metaboliltes in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation were performed during metabolites data preparation. The outcome of interest was standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there were `r nrow(WLM_train_dat)` remaining in the analysis. 5-Knot spline additive models for the gaussian outcome are constructed using two different models, the proposed BHAM-SSL and the SB-GAM. 10-Fold cross-validation(CV) are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples were used for prediction performance evaluation, where deviance, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean abosolute error (MAE)defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ were calculated and tabulated in Table \ref{tab:WLM_res}.

```{r}
rbind(
  'BHAM-SSL' = WLM_bamlasso_cv %>% slice_min(mse) %>% select(-s0),
  'SB-GAM' = WLM_SBGAM_cv %>%  slice_min(mse) %>% select(-lambda)
) %>% 
  rownames_to_column("Methods") %>% 
  flextable() %>% 
  flextable::autofit() %>% 
  flextable::set_caption(caption = "Measures of optimal BHAM-SSL and SB-GAM models for Weight Loss Maintenance Cohort by 10-fold cross-validation\\label{tab:WLM_res}")
```

### 2021-09-06 Manuscript/05-Discussion.Rmd
Manuscript revision

# Discussion
In the paper, we described a novel generalized additive model using Bayesian hierarchical priors, particularly the proposed spike-and-slab spline prior for bi-level functional selection. Meanwhile, we introduced two optimization based algorithms for model fitting. The algorithms can be easily scale up to address high-dimensional data analysis in a computational efficient manner, but also provide naive inferential supports for many commonly used hierarchical priors, e.g. t-distribution, double exponential distribution, mixture t distribution, and mixture double exponential distribution, in the analyses that the sparse assumption is weak or not necessary. Via simulations, we demonstrated that the proposed model provides as good, if not better, prediction performance compared to some state-of-the-art non-linear modeling devices.

>>>>

# Discussion
\label{sec:concl}
In the paper, we described a novel generalized additive model using Bayesian hierarchical priors, particularly the proposed spike-and-slab spline prior for bi-level functional selection. Meanwhile, we introduced two optimization based algorithms for model fitting. The algorithms can be easily scale up to address high-dimensional data analysis in a computational efficient manner, but also provide naive inferential supports for many commonly used hierarchical priors, e.g. t-distribution, double exponential distribution, mixture t distribution, and mixture double exponential distribution, in the analyses that the sparse assumption is weak or not necessary. Via simulations, we demonstrated that the proposed model provides as good, if not better, prediction performance compared to some state-of-the-art non-linear modeling devices.
